---
title: "Kaggle Covid19: Explorative Analysis"
author: "Florian Endel"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, cache=FALSE)

rm(list=ls())


## data
library(tidyverse)
library(lubridate)
#library(data.table)

## reading, writing
#library(readr)
#library(vroom)
#library(fst)
library(qs)
#library(arrow)
#library(feather)

## modeling
library(tidymodels)

## Document
library(knitr)
library(kableExtra)  # For nice tables
library(scales)      # For number formatting
library(cowplot)     # For arranging multiple plots

## utils
library(here)

## settings
theme_set(theme_light())


## source helper functions
source(here('src', 'directories_files.R'))
source(here('src', 'helper_functions.R'))



# Use this when you want to refit all models:
# clean_model_cache()


```


```{r data_kaggle_load}

kaggle <- qread(file = files$kaggle_interim)
model_data <- qread(file = files$kaggle_model_data)

```


```{r explorative_functions}


# Helper function for frequency tables
create_freq_table <- function(data, var_name) {
  data %>%
    count(!!sym(var_name), name = "Count") %>%
    drop_na() %>%
    mutate(
      Percentage = Count / sum(Count) * 100,
      Percentage = round(Percentage, 2)
    ) %>%
    kable(format = "html", caption = paste("Distribution of", var_name)) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
}

# Helper function for categorical variables visualization
plot_categorical <- function(data, var_name) {
  ggplot(data, aes(x = !!sym(var_name))) +
    geom_bar() +
    theme_minimal() +
    labs(title = paste("Distribution of", var_name),
         y = "Count") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Helper function for boolean variables
plot_boolean <- function(data, var_name) {
  ggplot(data, aes(x = !!sym(var_name), fill = !!sym(var_name))) +
    geom_bar() +
    scale_fill_manual(values = c("TRUE" = "#619CFF", "FALSE" = "#F8766D")) +
    theme_minimal() +
    labs(title = paste("Distribution of", var_name),
         y = "Count") +
    theme(legend.position = "none")
}

```


# Exploratory Data Analysis

```{r data_summary}
bool_columns <- c('PNEUMONIA', 'PREGNANT', 'DIABETES', 'COPD', 'ASTHMA', 'INMSUPR',
                 'HIPERTENSION', 'CARDIOVASCULAR', 'RENAL_CHRONIC', 'OTHER_DISEASE', 
                 'OBESITY', 'TOBACCO', 'INTUBED', 'ICU')

# Demographic Variables
## Age
age_summary <- kaggle %>%
  summarize(
    Mean = mean(AGE, na.rm = TRUE),
    Median = median(AGE, na.rm = TRUE),
    SD = sd(AGE, na.rm = TRUE),
    Min = min(AGE, na.rm = TRUE),
    Max = max(AGE, na.rm = TRUE)
  ) %>%
  kable(digits = 2) %>%
  kable_styling()

age_plot <- ggplot(kaggle, aes(x = AGE)) +
  geom_histogram(binwidth = 5, fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Age Distribution",
       x = "Age",
       y = "Count")

## Sex
sex_table <- create_freq_table(kaggle, "SEX")
sex_plot <- plot_categorical(kaggle, "SEX")

# Create plots for all boolean variables
bool_plots <- map(bool_columns, ~{
  plot_boolean(kaggle, .x)
})

# Arrange plots in a grid
bool_grid <- plot_grid(plotlist = bool_plots, ncol = 3)

# Patient Type
patient_type_table <- create_freq_table(kaggle, "PATIENT_TYPE")
patient_type_plot <- plot_categorical(kaggle, "PATIENT_TYPE")

# Classification Final
classification_table <- kaggle %>%
  count(CLASIFFICATION_FINAL) %>%
  mutate(Percentage = n/sum(n)*100) %>%
  kable() %>%
  kable_styling()

```


## Demographic Variables

### Age Distribution
```{r age-analysis}
age_summary
age_plot
```

### Sex Distribution
```{r sex-analysis}
sex_table
sex_plot
```

## Medical Conditions
```{r boolean-vars, fig.height=12, fig.width=12}
bool_grid
```

## Patient Management
```{r patient-type}
patient_type_table
patient_type_plot
```



## COVID-19 Classification

classification: covid test findings. Values 1-3 mean that the patient was diagnosed with covid in different degrees. 4 or higher means that the patient is not a carrier of covid or that the test is inconclusive.

```{r classification}
classification_table
```

```{r covid_classification_modified}
kaggle <- kaggle %>%
  mutate(
    covid_tested = CLASIFFICATION_FINAL <= 3,
    covid_tested_degree = case_when(
      CLASIFFICATION_FINAL == 1 ~ "1",
      CLASIFFICATION_FINAL == 2 ~ "2",
      CLASIFFICATION_FINAL == 3 ~ "3",
      TRUE ~ "no"
    ),
    covid_tested_degree = factor(covid_tested_degree, levels = c("1", "2", "3", "no"))
  )

# classification analysis
classification_analysis <- kaggle %>%
  summarize(
    total_cases = n(),
    tested_positive = sum(covid_tested),
    tested_negative = sum(!covid_tested),
    positive_rate = mean(covid_tested) * 100
  ) %>%
  mutate(
    across(where(is.numeric), ~round(., 2))
  ) %>%
  kable(col.names = c("Total Cases", "Tested Positive", "Tested Negative", "Positive Rate (%)"),
        caption = "Overall COVID-19 Testing Summary") %>%
  kable_styling()

# Detailed classification table
classification_detail <- kaggle %>%
  count(CLASIFFICATION_FINAL) %>%
  mutate(
    percentage = n/sum(n)*100,
    classification_meaning = case_when(
      CLASIFFICATION_FINAL == 1 ~ "COVID-19 Positive Degree 1",
      CLASIFFICATION_FINAL == 2 ~ "COVID-19 Positive Degree 2",
      CLASIFFICATION_FINAL == 3 ~ "COVID-19 Positive Degree 3",
      TRUE ~ "Negative or Inconclusive"
    )
  ) %>%
  arrange(CLASIFFICATION_FINAL) %>%
  kable(col.names = c("Classification", "Count", "Percentage", "Meaning"),
        digits = 2) %>%
  kable_styling()

# Visualization for covid_tested
covid_tested_plot <- ggplot(kaggle, aes(x = covid_tested, fill = covid_tested)) +
  geom_bar() +
  scale_fill_manual(values = c("TRUE" = "#619CFF", "FALSE" = "#F8766D")) +
  theme_minimal() +
  labs(title = "Distribution of COVID-19 Test Results",
       x = "Tested Positive",
       y = "Count") +
  theme(legend.position = "none")

# Visualization for covid_tested_degree
covid_degree_plot <- ggplot(kaggle, aes(x = covid_tested_degree, fill = covid_tested_degree)) +
  geom_bar() +
  scale_fill_manual(values = c("1" = "#619CFF", "2" = "#36A5FF", "3" = "#1E88E5", "no" = "#F8766D")) +
  theme_minimal() +
  labs(title = "Distribution of COVID-19 Test Degrees",
       x = "Test Degree",
       y = "Count") +
  theme(legend.position = "none")

# Combine plots
classification_plots <- plot_grid(covid_tested_plot, covid_degree_plot, 
                                ncol = 2, labels = "AUTO")

```


### Overall Testing Summary
```{r classification-summary}
classification_analysis
```

### Detailed Classification Distribution
```{r classification-detail}
classification_detail
```

### Visual Distribution of Test Results
```{r classification-plots}
classification_plots
```


