---
title: "Kaggle Covid19: Causal Analysis with DAGs"
author: "Florian Endel"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, cache=TRUE)

rm(list=ls())


## data
library(tidyverse)
library(lubridate)
#library(data.table)

## reading, writing
#library(readr)
#library(vroom)
#library(fst)
library(qs)
#library(arrow)
#library(feather)

## modeling
library(tidymodels)

## Document
library(knitr)
library(kableExtra)  # For nice tables
library(scales)      # For number formatting
library(cowplot)     # For arranging multiple plots

## utils
library(here)

## settings
theme_set(theme_light())


## source helper functions
source(here('src', 'directories_files.R'))
source(here('src', 'helper_functions.R'))


```

```{r setup_causal_dag, include=FALSE}
library(dagitty)
library(ggdag)
library(scales)
```


```{r data_kaggle_load}

kaggle <- qread(file = files$kaggle_interim)
model_data <- qread(file = files$kaggle_model_data)

```

# Causal Analysis with DAGs

In this section, we explore the causal relationships between different factors and COVID-19 outcomes. We use Directed Acyclic Graphs (DAGs) to formalize our causal assumptions and guide our analysis.

## DAG Specification and Initial Analysis

First, we create a DAG representing our hypothesized causal relationships. This graph shows how different medical conditions and demographic factors might influence COVID-19 mortality.

```{r setup_dag, message=FALSE, warning=FALSE}


# Create a DAG for our COVID outcomes
covid_dag <- dagitty('dag {
    Age -> Death
    Age -> Obesity
    Diabetes -> Death
    Diabetes -> Obesity
    Obesity -> Death
    Pneumonia -> Death
    Cardiovascular -> Death
    Hypertension -> Death
    Diabetes -> Hypertension
    Obesity -> Hypertension
    Age -> Hypertension
    Age -> Cardiovascular
    Obesity -> Cardiovascular
}')

# Visualize the DAG
ggdag(covid_dag, text_col='red', stylized = FALSE, label_col='red',
      text_size=2) + 
    theme_dag_blank() +
    labs(title = "Hypothesized Causal Relations for COVID-19 Mortality")
```


The DAG reveals several pathways through which obesity may affect mortality:
- Direct effect: Obesity -> Death
- Via cardiovascular system: Obesity -> Cardiovascular -> Death
- Via blood pressure: Obesity -> Hypertension -> Death
This multiple-pathway structure explains why controlling for mediators (like hypertension) is expected to reduce the total estimated effect.


## Identifying Causal Paths and Adjustment Sets

Next, we examine the paths between key variables and identify proper adjustment sets for causal estimation.

```{r dag_analysis_paths}
# Find all paths from Obesity to Death
paths_obesity_death <- paths(covid_dag, from = "Obesity", to = "Death")

# Create a nice table of paths
tibble(
  Path = paths_obesity_death$paths,
  `Open (d-connected)` = paths_obesity_death$open
) %>%
  kable(caption = "All Paths from Obesity to Death") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The table shows the paths from Obesity to Death, including whether the path is open (d-connected)^[d-connected: A path in a causal graph is considered d-connected (or "open") if it transmits association (correlation). This means that the variables in the path are statistically related unless they are blocked by a collider. We are not adjusting for colliders in the given example.]

```{r dag_analysis_adjustment}
# Identify minimal adjustment sets
adj_sets <- adjustmentSets(covid_dag, exposure = "Obesity", outcome = "Death")

# Create a nice table of adjustment sets
tibble(
  `Adjustment Set` = sapply(adj_sets, paste, collapse = ", ")
) %>%
  kable(caption = "Minimal Adjustment Sets for Estimating Obesity's Effect on Death") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


The DAG analysis identifies that we need to adjust for **Age** and **Diabetes** to estimate the causal effect of obesity on mortality. This means:

1. Age and diabetes are confounders - they affect both obesity and death risk independently
2. Without adjusting for these variables, our estimate of obesity's effect on death would be biased
3. After adjusting for age and diabetes, any remaining association between obesity and death can be interpreted more causally

This suggests that in our statistical analyses, we should always control for age and diabetes when estimating the effect of obesity on COVID-19 mortality. Other variables like hypertension and cardiovascular disease are potential mediators (they lie on the causal pathway between obesity and death) and should not be adjusted for if we want to estimate the total causal effect of obesity.

```{r adjustment_sets_visual}
# Visualize the adjustment set
ggdag_adjustment_set(covid_dag,
                    exposure = "Obesity",
                    outcome = "Death",
                    text_col = "red",
                    stylized = FALSE,
                    text_size = 2) +
    theme_dag_blank() +
    labs(title = "Minimal Adjustment Set for Obesity -> Death Effect")
```

The visualization highlights our minimal adjustment set (Age and Diabetes) in red. These variables must be controlled for to estimate the causal effect of obesity on mortality. The red paths show the relationship we're interested in (from Obesity to Death), while other paths represent various causal pathways in our model. Variables like Hypertension and Cardiovascular disease are shown in their original color as they are mediators and should not be adjusted for when estimating the total causal effect.

## Testing DAG Implications with Data

Now we'll test whether our data is consistent with the causal assumptions encoded in our DAG.

```{r data_prep}
# Prepare analysis dataset
analysis_data <- model_data %>%
  transmute(
    Death = as.logical(DIED),
    Age = as.integer(AGE),
    Diabetes = as.integer(DIABETES),
    Obesity = as.integer(OBESITY),
    Pneumonia = as.integer(PNEUMONIA),
    Cardiovascular = as.integer(CARDIOVASCULAR),
    Hypertension = as.integer(HIPERTENSION)
  )

# Function to test conditional independence
test_conditional_independence <- function(data, var1, var2, conditioning) {
  formula <- as.formula(paste(var1, "~", var2, "+", 
                            paste(conditioning, collapse = "+")))
  model <- glm(formula, data = data, family = binomial())
  coef_summary <- summary(model)$coefficients
  tibble(
    Variable = var2,
    Estimate = coef_summary[var2, "Estimate"],
    `P-value` = coef_summary[var2, "Pr(>|z|)"]
  )
}

# Test some key relationships
independence_tests <- bind_rows(
  test_conditional_independence(analysis_data, "Death", "Obesity", 
                              c("Age", "Diabetes")),
  test_conditional_independence(analysis_data, "Death", "Hypertension", 
                              c("Age", "Obesity"))
)

independence_tests %>%
  kable(caption = "Tests of Conditional Independence",
        digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The test results show significant associations (p < 0.001) between:

1. Obesity and Death, even after adjusting for Age and Diabetes, supporting a causal relationship
2. Hypertension and Death, even after adjusting for Age and Obesity, suggesting an independent effect

These results are consistent with our DAG's structure.


## Estimating Causal Effects

Based on the adjustment sets identified by our DAG, we estimate the causal effect of obesity on mortality.

```{r causal_effects_models}
# Model with minimal adjustment set
model_dag <- glm(Death ~ Obesity + Age + Diabetes,  # minimal adjustment set
                 data = analysis_data, 
                 family = binomial())

# For comparison: Inappropriate model with mediator
model_mediator <- glm(Death ~ Obesity + Age + Diabetes + Hypertension,  # including mediator
                      data = analysis_data, 
                      family = binomial())
```

```{r causal_effects_results}
# Compare results
bind_rows(
  broom::tidy(model_dag) %>% 
    mutate(model = "DAG-based (correct adjustment)"),
  broom::tidy(model_mediator) %>% 
    mutate(model = "With mediator (over-adjusted)")
) %>%
  filter(term == "Obesity") %>%
  select(model, estimate, std.error, p.value) %>%
  mutate(
    OR = exp(estimate),
    CI_lower = exp(confint(model_dag)["Obesity", 1]),
    CI_upper = exp(confint(model_dag)["Obesity", 2])
  ) %>%
  kable(caption = "Estimated Causal Effect of Obesity on Mortality",
        digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Interpretation

The DAG-based analysis shows that obesity has a significant causal effect on COVID-19 mortality. The properly adjusted model (controlling for age and diabetes) indicates that obesity increases the odds of death by `r scales::percent(exp(coef(model_dag)["Obesity"])-1,accuracy=0.01)` (OR = `r round(exp(coef(model_dag)["Obesity"]), 2)`, 95% CI: `r round(exp(confint(model_dag)["Obesity", 1]), 2)`-`r round(exp(confint(model_dag)["Obesity", 2]), 2)`).

When we additionally control for hypertension (a mediator), the estimated effect decreases to `r scales::percent(exp(coef(model_mediator)["Obesity"])-1,accuracy=0.01)` (OR = `r round(exp(coef(model_mediator)["Obesity"]), 2)`, 95% CI: `r round(exp(confint(model_mediator)["Obesity", 1]), 2)`-`r round(exp(confint(model_mediator)["Obesity", 2]), 2)`). This reduction is expected because we're blocking one of the pathways through which obesity affects mortality (via hypertension), thus capturing only part of obesity's total causal effect.

This analysis suggests that obesity has both direct effects on COVID-19 mortality and indirect effects through conditions like hypertension. For public health purposes, the larger total effect (`r scales::percent(exp(coef(model_dag)["Obesity"])-1,accuracy=0.01)` increased odds) is most relevant, as it captures the full impact of obesity on COVID-19 mortality risk.


The DAG analysis reveals several important insights:

1. Obesity affects mortality both directly and through multiple indirect paths (via Hypertension and Cardiovascular disease).
2. Age is a key confounder that needs to be adjusted for in any analysis.
3. Different adjustment sets lead to somewhat different effect estimates, highlighting the importance of careful consideration of causal pathways.

## Limitations

This causal analysis has several limitations:
- The DAG represents our assumptions about causal relationships but cannot prove causality.
- We assume no unmeasured confounding.
- The relationships are likely more complex than shown in our simplified DAG.
- Missing data and measurement error could bias our estimates.

